Towards Practical First-Order Model Counting

### 2024-04-25 doc/paper.tex
Section 2.2

\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Recently, a new algorithm for FOMC
  was proposed that, instead of simply providing the final count, generates
  definitions of (possibly recursive) functions, which can be evaluated with
  different arguments to compute the model count for any domain size. However,
  the algorithm did not include base cases in the recursive definitions. This
  work makes three contributions. Firstly, we demonstrate how to construct
  function definitions that include base cases by modifying the logical formulas
  used in the FOMC algorithm. Secondly, we extend the well-known circuit
  modification technique in knowledge compilation, known as smoothing, to work
  with the formulas corresponding to base cases and the recently proposed node
  types, such as domain recursion and constraint removal. Thirdly, we introduce
  a compilation algorithm that transforms the function definitions into C++
  code, equipped with arbitrary-precision arithmetic. These additions allow the
  new FOMC algorithm to scale to domain sizes over \num{9000} times larger than
  the current state of the art, as demonstrated through experimental results.
\end{abstract}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item 9 pages including appendices but not acknowledgements
    \item Add some papers mentioned in:
          \begin{itemize}
            \item very recent work
            \item my previous paper, including:
                  \begin{itemize}
                    \item other liftable fragments
                    \item some more theory papers, e.g., LICS 2018
                  \end{itemize}
          \end{itemize}
  \end{itemize}
}

@@

\subsection{Logic (TODO)}\label{sec:logic}

@@

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For every predicate $p/n$ in $\phi$, let
  ${(\Delta_{i}^{p})}_{i=1}^{n}$ be a list of the corresponding domains (not
  necessarily distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets such that:
  \begin{itemize}
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{itemize}
  Then a \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of
  ground literals defined by adding either $p(\mathbf{t})$ or
  $\neg p(\mathbf{t})$ for every predicate $p/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{p})$. A structure is a
  \emph{model} if it satisfies $\phi$ (see Appendix~A of~Dilkas and
  Belle~\shortcite{DBLP:journals/corr/abs-2306-04189} for more details).
\end{definition}

\begin{definition}[WFOMC]
  Continuing from \cref{def:model}, for every predicate $p/n$ in $\phi$, let
  $w^{+}(p), w^{-}(p) \in \mathbb{R}$ be its (positive and negative)
  \emph{weights}. Unless explicitly specified otherwise, we assume weights to be
  equal to one. The \emph{(symmetric) weighted first-order model count} (WFOMC)
  of $\phi$ (with respect to $\sigma$, $w^{+}$, and $w^{-}$) is the quantity
  \[
    \sum_{M \models \phi} \prod_{p(\mathbf{t}) \in M} w^{+}(p) \prod_{\neg p(\mathbf{t}) \in M} w^{-}(p),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

@@

\paragraph{Three Types of Logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item In many-sorted logic, each term is assigned to a \emph{sort}, and each
        predicate $p/n$ is assigned to a sequence of $n$ sorts. Each sort has
        its corresponding domain. In the input formula, all domains are assumed
        to be pairwise disjoint Most of these assignments are typically left
        implicit and can be reconstructed from the quantifiers. For instance,
        $\forall x,y \in \Delta\text{. }P(x, y)$ implies that variables $x$ and
        $y$ have the same sort. On the other hand, $\forall x \in \Delta\text{.
        }\forall y \in \Gamma\text{. } P(x, y)$ implies that $x$ and $y$ have
        different sorts, and it would be improper to have $x = y$ as part of a
        formula.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

\begin{example}\label{example:functions}
  Functions
  \begin{itemize}
    \item In \Ctwo: $\forall x \in \Delta\text{.
          }\exists^{=1} y \in \Delta\text{. }P(x, y)$
    \item In \UFO:
          \begin{gather*}
            \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
            |P| = |\Delta|
          \end{gather*}
    \item In \FO:
    \begin{equation}\label[formula]{eq:fo}
      \begin{gathered}
        \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
        \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z
      \end{gathered}
    \end{equation}
  \end{itemize}

  Note that \cref{eq:fo} is written with two domains instead of one. Clearly,
  the model count does not change whenever $|\Gamma| = |\Delta|$. As was
  previously observed by Dilkas and Belle~\shortcite{DBLP:conf/kr/DilkasB23},
  such a reformulation can help first-order knowledge compilation algorithms
  find efficient solutions.
\end{example}

@@

\clearpage % TODO: temporarary

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

@@

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Recently, a new algorithm for FOMC
  was proposed that, instead of simply providing the final count, generates
  definitions of (possibly recursive) functions, which can be evaluated with
  different arguments to compute the model count for any domain size. However,
  the algorithm did not include base cases in the recursive definitions. This
  work makes three contributions. First, we demonstrate how to construct
  function definitions that include base cases by modifying the logical formulas
  used in the FOMC algorithm. Second, we extend the well-known circuit
  modification technique in knowledge compilation, known as smoothing, to work
  with the formulas corresponding to base cases and the recently proposed node
  types, such as domain recursion and constraint removal. Third, we introduce a
  compilation algorithm that transforms the function definitions into C++ code,
  equipped with arbitrary-precision arithmetic. These additions allow the new
  FOMC algorithm to scale to domain sizes over \num{9000} times larger than the
  current state of the art, as demonstrated through experimental results.
\end{abstract}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item 9 pages including appendices but not acknowledgements
    \item Add some papers mentioned in:
          \begin{itemize}
            \item very recent work
            \item my previous paper, including:
                  \begin{itemize}
                    \item other liftable fragments
                    \item some more theory papers, e.g., LICS 2018
                  \end{itemize}
          \end{itemize}
    \item introduce \Cranetwo{}
    \item (maybe) introduce lifted/liftable
  \end{itemize}
}

@@

\todo[inline]{There are three subsections now, not two.}

\subsection{First-Order Logic and WFOMC (TODO)}\label{sec:logic}

@@

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For every predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains (not
  necessarily distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets such that:
  \begin{enumerate*}[label=(\roman*)]
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate*}
  Then a \emph{structure} of $\phi$ (with respect to $\sigma$) is a set $M$ of
  ground literals defined by adding either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$ (see Appendix~A of~Dilkas and
  Belle~\shortcite{DBLP:journals/corr/abs-2306-04189} for more details).
\end{definition}

\begin{definition}[WFOMC]
  Continuing from \cref{def:model}, for every predicate $P/n$ in $\phi$, let
  $w^{+}(P), w^{-}(P) \in \mathbb{R}$ be its (positive and negative)
  \emph{weights}. Unless specified otherwise, we assume weights to be equal to
  one. The \emph{(symmetric) weighted first-order model count} (WFOMC) of $\phi$
  (with respect to $\sigma$, $w^{+}$, and $w^{-}$) is the quantity
  \[
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

\subsection{WFOMC Algorithms and Their Logics}\label{sec:threelogics}

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in WFOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{},
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}, and
  \item \UFO{} is the input format supported by a private version of
  \textsc{FastWFOMC} obtained directly from the authors.
\end{enumerate*}
Note that the publicly available
version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
does not support any cardinality constraints. The notation we use to refer to
each logic is standard in the case of \Ctwo{}, new in the case of \UFO{}, and
redefined to be more specific in the case of \FO{}. All three logics are
function-free, and domains are always assumed to be finite.

In \FO{}, each term is assigned to a \emph{sort}, and each predicate $P/n$ is
assigned to a sequence of $n$ sorts. Each sort has its corresponding domain.
Most of these assignments to sorts are typically left implicit and can be
reconstructed from the quantifiers. For example, $\forall x,y \in \Delta\text{.
}P(x, y)$ implies that variables $x$ and $y$ have the same sort. On the other
hand, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$
implies that $x$ and $y$ have different sorts, and it would be improper to
write, e.g., $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
clauses with more than two variables, and the equality predicate.

\begin{remark}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm will be able to compile the
  formula into a circuit or graph suitable for lifted model counting. However,
  it is known that \textsc{ForcLift} compilation is guaranteed to succeed on any
  \FO{} formula without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11}.
\end{remark}

Both \Ctwo{} and \UFO{} are single-sorted (i.e., variables are always quantified
over the same domain) and only include formulas with at most two variables and
no constants. Unlike \UFO{}, \Ctwo{} supports existential quantifiers, including
\emph{counting quantifiers}. For example, $\exists^{=1} x\text{. }\phi(x)$ means
that there exists \emph{exactly one} $x$ such that $\phi(x)$, and
$\exists^{\le 2} x\text{. }\phi(x)$ means that there exist \emph{at most two}
such $x$. The advantage \UFO{} brings over \Ctwo{} is the support for
\emph{(equality) cardinality constraints}. For example, $|P| = 3$ constrains all
models to have \emph{exactly three positive literals with predicate $P$}. We
further illustrate the differences among the three logics with the example
below.

\begin{example}[Functions]\label{example:functions}
  To define predicate $P$ to be an endofunction on $\Delta$, in \Ctwo{} one
  could write $\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{.
  }P(x, y)$. In \UFO{}, the same could be written as
  \begin{equation}\label[formula]{eq:functions1}
    \begin{gathered}
      \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
      |P| = |\Delta|.
    \end{gathered}
  \end{equation}
  Although \cref{eq:functions1} has some additional models, the negative weight
  $w^{-}(S) = -1$ makes some of the counts cancel out so that the WFOMC of
  \cref{eq:functions1} is still equal to the number of endofunctions on
  $\Delta$.

  In \FO{}, we would write
  \begin{equation}\label[formula]{eq:fo}
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z,
    \end{gathered}
  \end{equation}
  first stating that each $x$ must be mapped to at least one $y$ and then adding
  that if $x$ is mapped to both $y$ and $z$, then $y=z$. Note that \cref{eq:fo}
  is written with two domains instead of one. Clearly, one can still count the
  number of endofunctions by setting $|\Gamma| = |\Delta|$. As was previously
  observed by Dilkas and Belle~\shortcite{DBLP:conf/kr/DilkasB23}, such a
  reformulation can help first-order knowledge compilation algorithms find
  efficient solutions.
\end{example}

@@
